---
title: "Homework7"
author: "Chen Liu"
date: "2/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the options data from file OptionsPrices.csv used in the workshop calculate the P&L breakdown into delta, gamma and vega components of 1425 December 2007 put for each day between 2/27/2007 and 3/5/2007.
Compare actual P&L for 1425 December 2007 put with the estimated daily P&L for each day between 2/27/2007 and 3/5/2007.
Put the results in table with columns: actual P&L, estimated P&L, delta P&L, gamma P&L, vega P&L.

```{r,include=FALSE}
library(tidyverse)
library(RQuantLib)
```

```{r}
datapath<-"/Users/me/Desktop/MSCA/FinancialAnalytics/week7"
optionsData <- read.csv(paste(datapath,'OptionsPrices.csv',sep='/'),header=TRUE,row.names=1,sep=",")
```

### 1. calculate the P&L breakdown into delta, gamma and vega components of 1425 December 2007 put for each day between 2/27/2007 and 3/5/2007.

#### Create function implying volatility structure from put prices of one row of optionsData.

```{r}
# Calculate implied volatility for puts
impVolPuts<-apply(as.matrix(optionsData),1,
                      function(z) EuropeanOptionImpliedVolatility("put", 
                                                                  value=z[26],
                                                                  underlying = z[3],
                                                                  strike=1425,
                                                                  dividendYield = z[2],
                                                                  riskFreeRate = z[2],
                                                                  maturity = z[1]/365,
                                                                  volatility=0.3))

head(impVolPuts)

```

```{r}
n <-nrow(optionsData)
#Calculate option value,delta,gamma,vega
value<-rep(NA,n) 
delta<-rep(NA,n) 
gamma<-rep(NA,n) 
vega<-rep(NA,n) 

for (i in 1:n){
  value[i] <- unlist(EuropeanOption(type="put", 
                                  underlying=optionsData[i,3], 
                                  strike=1425,
                                  dividendYield=optionsData[i,2],
                                  riskFreeRate=optionsData[i,2],
                                  maturity=optionsData[i,1]/365, 
                                  volatility=impVolPuts[i]))[1] 
  delta[i] <- unlist(EuropeanOption(type="put", 
                                  underlying=optionsData[i,3], 
                                  strike=1425,
                                  dividendYield=optionsData[i,2],
                                  riskFreeRate=optionsData[i,2],
                                  maturity=optionsData[i,1]/365, 
                                  volatility=impVolPuts[i]))[2] 
  gamma[i] <- unlist(unlist(EuropeanOption(type="put", 
                                  underlying=optionsData[i,3], 
                                  strike=1425,
                                  dividendYield=optionsData[i,2],
                                  riskFreeRate=optionsData[i,2],
                                  maturity=optionsData[i,1]/365, 
                                  volatility=impVolPuts[i]))[3]) 
  vega[i] <- unlist(unlist(EuropeanOption(type="put", 
                                  underlying=optionsData[i,3], 
                                  strike=1425,
                                  dividendYield=optionsData[i,2],
                                  riskFreeRate=optionsData[i,2],
                                  maturity=optionsData[i,1]/365, 
                                  volatility=impVolPuts[i]))[4]) 
}
summary2<-data.frame('Value'=value,'Delta'= delta,'Gamma'=gamma,'Vega'=vega) 
rownames(summary2) <- rownames(optionsData)
head(summary2)
```

```{r}

PutPNL.Delta<-rep(NA,209) 
PutPNL.Gamma<-rep(NA,209) 
PutPNL.Vega<-rep(NA,209)
for (i in 2:n) {
  PutPNL.Delta[i]<-delta[i-1]*(optionsData[i,3]-optionsData[i-1,3])
  PutPNL.Gamma[i]<-.5*gamma[i-1]*(optionsData[i,3]-optionsData[i-1,3])^2
  PutPNL.Vega[i]<-vega[i-1]*
                      (impVolPuts[i]-impVolPuts[i-1])
}
putPNL.estimated<-data.frame(delta=PutPNL.Delta,gamma=PutPNL.Gamma,vega=PutPNL.Vega)
rownames(putPNL.estimated)<-rownames(optionsData)
head(putPNL.estimated)
```

```{r}
putPNL.estimatedTotal<-apply(putPNL.estimated,1,sum)
putPNL.actual<-c(NA,diff(optionsData$P1425))
```

## Create table
```{r}
PNL<-data.frame(actual=putPNL.actual,estimated=putPNL.estimatedTotal,
                delta=PutPNL.Delta,gamma=PutPNL.Gamma,vega=PutPNL.Vega)
rownames(PNL)<-rownames(optionsData)
head(PNL)
```



































