---
title: "FA_project"
author: "Chen Liu"
date: "3/15/2020"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Course Project: Step1
# Efficient Frontier and CAPM, Market-Neutral Portfolios

## 1. Description of Portfolio
Read portfolio consisting of 297 S&P 500 stocks prices for 2014.
The data are in the file PortfolioSP500Stocks.csv.
In addition to stock prices the file also contains prices of S&P 500 SPDR ETF SPY and Fed Funds rates for the same period.
```{r,load data}
datapath<-'/Users/me/Desktop/MSCA/FinancialAnalytics/project'
Data2014<-read.csv(paste(datapath,'PortfolioSP500Stocks.csv',sep='/'),header=TRUE)
dim(Data2014)
head(Data2014[,1])
colnames(Data2014)
```

```{r}
#Transform dates in the first column into date format.
Data2014[,1]<-as.Date(Data2014[,1],origin = "1899-12-30")
head(Data2014[,1:3])
```

## 2. Efficient Frontier
Calculate means and standard deviations of daily log-returns for each company.
Calculate mean and standard deviation of daily log-returns for SPY.
Calculate mean Fed Funds rate for 2014.
```{r}
Mean.Sd.Data2014<-cbind(sd=apply(Data2014[,-c(1,3)],2,function(z) sd(diff(log(z)),na.rm=TRUE)),
                        mean=apply(Data2014[,-c(1,3)],2,function(z) mean(diff(log(z)),na.rm=TRUE)))
head(Mean.Sd.Data2014)
Mean.FedFunds<-mean(Data2014[,3])/100/250
```
Plot the SPY companies on standard deviation-mean diagram.
Observe efficient frontier.
Add the points for SPY and risk-free rate.
Add the Capital Allocation Line connecting the points of risk-free rate and SPY.
```{r}
plot(Mean.Sd.Data2014[-1,],ylab="Mean",xlab="Sd",pch=19,xlim=c(0,.04),ylim=c(-.0040,.004))
points(Mean.Sd.Data2014[1,1],Mean.Sd.Data2014[1,2],col="red",pch=19)
points(0,Mean.FedFunds,col="green",pch=19)# Plot the risk free return in green
lines(c(0,4*Mean.Sd.Data2014[1,1]),c(Mean.FedFunds,4*Mean.Sd.Data2014[1,2]),col="red")# Draw a line that contains both the Risk free return and the SPY
text(Mean.Sd.Data2014[1,1],Mean.Sd.Data2014[1,2],labels="SPY",cex=.8,col="red",pos=2)
abline(v=Mean.Sd.Data2014[1,1])
```

*Why is the point representing SPY located at the tip of the frontier?*
Because the SPY represents the collection of all the individual stocks. It has the smallest standard deviation since the movement of one stock will only fractionally affect its total return (1 out of 500 stocks). Correspondingly, an individual stock will have much more variance in its daily returns. Some more than others. 
If you consider where the SPY is on the Y axis. It is at the value that represents the mean of all the stocks, the mean value of all the individual stock's means(on a cap-weighted basis).

Find the best capital allocation line in the portfolio, i.e. the tangent to the efficient frontier.
Find the stock name forming the tangent portfolio (call it XYZ for now).
```{r}
which.min(Mean.Sd.Data2014[Mean.Sd.Data2014[,2]>.002,][,1]) #KR
# Print SD and mean of KR
Mean.Sd.Data2014["KR",]
```
```{r}
plot(Mean.Sd.Data2014[-1,],ylab="Mean",xlab="Sd",pch=19,xlim=c(0,.04),ylim=c(-.0040,.004))
points(Mean.Sd.Data2014[1,1],Mean.Sd.Data2014[1,2],col="red",pch=19)
points(0,Mean.FedFunds,col="green",pch=19)
lines(c(0,4*Mean.Sd.Data2014[1,1]),c(Mean.FedFunds,4*Mean.Sd.Data2014[1,2]),col="red")
text(Mean.Sd.Data2014[1,1],Mean.Sd.Data2014[1,2],labels="SPY",cex=.8,col="red",pos=2)
abline(v=Mean.Sd.Data2014[1,1])
# Add a line that connects the risk free rate and the Mean/SD for KR
lines(c(0, 3*Mean.Sd.Data2014["KR",1]), c(Mean.FedFunds,3*Mean.Sd.Data2014["KR",2]), col="orange")
# Label KR
text(Mean.Sd.Data2014[131,1], Mean.Sd.Data2014[131,2], labels="KR", cex=1, col="orange", pos=2)

```
Plot cumulative returns of SPY and XYZ
```{r}
# Create a vector that contains the month/day of each observation from the original data
labDates <- paste(format(Data2014[,1],  "%m"), format(Data2014[,1], "%d"),sep="/")
#Plot the cumulative sum of the log differences of SPY and of KR
matplot(Data2014$Date[-1], cbind(cumsum(diff(log(Data2014$SPY))),(cumsum(diff(log(Data2014$KR))))), type="l",lty=1,lwd=3,col=c("red","gold"),
        ylab="Cumulative Returns",xlab="Date",xaxt="n")
axis(side=1,labels=labDates[-1],at=Data2014$Date[-1])
legend("topleft",legend=c("SPY","XYZ"),lty=1,col=c("red","gold"))
```
Create portfolio of the risk free investment and XYZ that has the same risk as SPY, but with better return.
The portfolio point on the efficient frontier chart should be on the capital allocation line for XYZ.

```{r}
# Ratio of standard deviation of SPY  and KR is the weight
portfolioWeight<-Mean.Sd.Data2014[1,1]/Mean.Sd.Data2014["KR",1]
# Mix of weighted cumulative sums of log differenced Risk free rate and KR 
mixRFKR<-(1-portfolioWeight)*cumsum(rep(Mean.FedFunds,length(Data2014[,1])))+
          portfolioWeight*cumsum(c(0,diff(log(Data2014$KR))))
# Calculate the daily returns of the mixed portfoloio of Risk free rate and KR for 250 days
mixRFKR.returns<-(1-portfolioWeight)*rep(Mean.FedFunds,length(Data2014[,1]))+
          portfolioWeight*c(0,diff(log(Data2014$KR)))
# Cumulative sum of the mixed portfolio's daily returns for 250 days
mixRFKR.cum.returns<-cumsum(mixRFKR.returns)
# Plot the cumulative sums of returns of SPY, KR and the mixed portfolio
matplot(Data2014[,1],
        cbind(cumsum(c(0,diff(log(Data2014$SPY)))),
              cumsum(c(0,diff(log(Data2014$KR)))),
              mixRFKR.cum.returns),type="l",lty=1,lwd=3,col=c("red","gold","purple"),
        ylab="Cumulative Returns",xlab="Date",xaxt="n")
# label the dates
axis(side=1,labels=labDates,at=Data2014[,1])
# add the legend
legend("topleft",legend=c("SPY","XYZ","Mixed Portfolio"),lty=1,lwd=3,col=c("red","gold","purple"))
```

Check that risk of the mixed portfolio is the same as risk of SPY.
```{r}
c(RiskSPY=sd(diff(log(Data2014$SPY))),RiskMixedPortfolio=sd(mixRFKR.returns))
```

## CAPM
Define the CAPM model for the portfolio and plot the Security Market Line.
```{r}
# De-annualize the fed funds rates
FedFunds.daily<-Data2014$FFRate/100/365 
# Run linear regression and extract the coefficients
portfolioBetas<-apply(as.matrix(Data2014[,-c(1:3)]),2,
                      function(z) lm(I(diff(log(z))-FedFunds.daily[-length(FedFunds.daily)])~-1+                                     I(diff(log(Data2014$SPY))-FedFunds.daily[-length(FedFunds.daily)]))$coefficients)
#Plot the Betas for each stock in the data set
plot(1:length(portfolioBetas),portfolioBetas,xlab="Index")
```
Betas are estimated as slopes of each excess log returns sample to the SPY excess log returns.

Select stocks Coca-Cola (KO), Plum Creek Timber (PCL) and Kroger (KR).
Find their betas.
```{r}
# select the indexed values from the list of stocks
selectedStocksIndex.x<-c(match("KO",names(portfolioBetas)),
                       match("PCL",names(portfolioBetas)),
                       match("KR",names(portfolioBetas)))
# select the 3 stocks from portfolio Betas
(selectedStocks.x<-portfolioBetas[selectedStocksIndex.x])
# Select the indexed values from 
selectedStocksIndex.y<-c(match("KO",names(Mean.Sd.Data2014[,2])),
                       match("PCL",names(Mean.Sd.Data2014[,2])),
                       match("KR",names(Mean.Sd.Data2014[,2])))
# Calculate the mean return of KO, PCL, KR
(selectedStocks.y<-Mean.Sd.Data2014[,2][selectedStocksIndex.y])
```
Create plot in the Betas-Mean Return space:

  * Put all stocks of the portfolio on the graph
  * Add SPY, the risk free asset and the SML to the plot
  * Mark the selected stocks
```{r}
#Assign SPY Beta to 1
SPY.Beta <- 1
# Plot all the Betas and Mean returns
plot(portfolioBetas, Mean.Sd.Data2014[-1,2],ylab="Mean",xlab="Portfolio Betas",main="Stock Mean vs Beta", pch=19,xlim=c(0,2))
# Plot SPY point
points(SPY.Beta ,Mean.Sd.Data2014[1,2],col="red",pch=19)
# Label the SPY point
text(SPY.Beta ,Mean.Sd.Data2014[1,2],labels="SPY",cex=.8,col="red",pos=1)
# Plot Risk free return point
points(0,Mean.FedFunds,col="green",pch=19)
# Draw the Security Market Line
lines(c(0,2*1),c(Mean.FedFunds,2*Mean.Sd.Data2014[1,2]),col="red")
# Plot and label the 3 stocks KO, PCL, KR
points(selectedStocks.x[1],selectedStocks.y[1], col="gold", pch=16)
text(selectedStocks.x[1],selectedStocks.y[1], labels="KO", cex=.8, pos=2)
points(selectedStocks.x[2],selectedStocks.y[2], col="gold", pch=16)
text(selectedStocks.x[2],selectedStocks.y[2], labels="PCL", cex=.8, pos=2)
points(selectedStocks.x[3],selectedStocks.y[3], col="gold", pch=16)
text(selectedStocks.x[3],selectedStocks.y[3], labels="KR", cex=.8, pos=2)

```
*According to CAPM model which of the selected stocks should have been bought in 2014 and which should have been shorted?*
Because PCL is below the SML, this company is considered overvalued in price and should have been shorted.KR is above the SML and is undervalued, KR should have been bought in 2014.

## Market-Neutral Portfolio
A portfolio is called long only if it consists of only long positions on assets.
Similarly, portfolio is short only if it consists only of short positions on assets.
Portfolio is called market-neutral if it has both long and short groups of assets and the initial dollar values of these two groups are equal or periodically maintained equal.

Create market-neutral long-short portfolio of the 3 selected stocks?
Run back test of the created portfolio.
```{r}
c(KR.price=Data2014$KR[1],PCL.price=Data2014$PCL[1])
longWeight<-Data2014$PCL[1]/Data2014$KR[1]
longWeight
```

Long weight shows that for each shorted share of PCL the market-neutral portfolio should buy 2.24 shares of KR.
```{r}
portfolioValue<-longWeight*Data2014$KR-Data2014$PCL
# find the price ratio of the 2 stocks based on the first day's prices
plot(portfolioValue,type="l",xlab="2014",ylab="Value of Market-Neutral Portfolio")
```

Create market-neutral portfolio of stocks according to the CAPM as of the beginning of 2014 and track its value for the rest of the year.

Define the equation parameters of SML.
The two points on this line are: <x1,y1>=<0,Mean.FedFunds> and <x2,y2>=<1,Mean.Sd.Data2014[1,2]<.
Then the line equation is y(x)=a+bx, where a=y1= Mean.FedFunds and b=y(x2)−ax2= Mean.Sd.Data2014[2]-Mean.FedFunds.
```{r}
SML<-data.frame(a=Mean.FedFunds,b=Mean.Sd.Data2014[1,2]-Mean.FedFunds)
#Long only
```
Create and plot long only portfolio.
```{r}
# Select those stocks whose Mean is greater than the mean of the SML for a given value of Beta
longPortfolio<-Mean.Sd.Data2014[-1,2]>(SML$a+SML$b*portfolioBetas)
# Print the names of the stocks for the long portfolio
rownames(Mean.Sd.Data2014[-1,])[longPortfolio]
```

Plot the division of stocks by the SML

```{r}
# Plot all the points
plot(portfolioBetas,Mean.Sd.Data2014[-1,2],ylab="Mean Returns",xlab="Betas",pch=19,xlim=c(0,2))
# Plot the SPY point
points(1,Mean.Sd.Data2014[1,2],col="red",pch=19)
# Plot the Risk Free Rate point
points(0,Mean.FedFunds,col="green",pch=19)
# Draw the SML line
lines(c(0,2),c(Mean.FedFunds,2*Mean.Sd.Data2014[1,2]),col="red")
# Highlight the long portfolio (stocks above SML)
points(portfolioBetas[longPortfolio],Mean.Sd.Data2014[-1,2][longPortfolio],col="gold",pch=16)
```

Calculate weights of the long only portfolio based on the distance to MAL.
If di=μi−(a+b∗βi) then the portfolio weights wi=di∑ni=1di.
```{r}
# calculate the distance from the mean of each stock to the SML
longOnlyDistance.d<-Mean.Sd.Data2014[-1,2][longPortfolio]-(SML$a+SML$b*portfolioBetas[longPortfolio])
# Calculate the weights of each stock based on each stock's distance
longOnlyWeights<-longOnlyDistance.d/sum(longOnlyDistance.d)
head(longOnlyWeights)
plot(longOnlyWeights,type="b")
# Confirm that all the weights sum to 1
sum(longOnlyWeights)
```
Calculate the initial value of weighted long portfolio.
```{r}
longOnlyValue<-as.matrix(Data2014[1,-(1:3)])[longPortfolio]%*%longOnlyWeights
```
Create short portfolio
```{r}
# Calculate the distance of each short stock mean to the SML
shortOnlyDistance.d<-(SML$a+SML$b*portfolioBetas[!longPortfolio])-Mean.Sd.Data2014[-1,2][!longPortfolio]
# Use the percentage of individual distance as a weight, for each stock
shortOnlyWeights<-shortOnlyDistance.d/sum(shortOnlyDistance.d)
head(shortOnlyWeights)
plot(shortOnlyWeights,type="b")
sum(shortOnlyWeights)
```

```{r}
#Calculate the initial value of weighted short portfolio.
shortOnlyValue<-as.matrix(Data2014[1,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights
#Find the proportion between the long and the short portfolio.
c(longOnlyValue=longOnlyValue,shortOnlyValue=shortOnlyValue)
portfolioProportion<-shortOnlyValue/longOnlyValue
unclass(portfolioProportion)
c(longOnlyShares=shortOnlyValue/longOnlyValue,shortOnlyShares=1)
```
Calculate value trajectory of the total portfolio and plot it.
```{r}
longValueTrajectory<-as.matrix(Data2014[,-(1:3)])[,longPortfolio]%*%longOnlyWeights
shortValueTrajectory<-as.matrix(Data2014[,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights
totalPortfolioTrajectory<-longValueTrajectory%*%portfolioProportion-shortValueTrajectory
plot(totalPortfolioTrajectory,type="l",xlab="2014",ylab="Value of Market-Neutral Portfolio")

```
*How would you calculate the annual return of this portfolio?*
The formula to calculate the return equals: (final value-initial value)/initial value
The value of this portfolio starts at 0 and ends at about 40. But we can not use 0 as the initial value because there is initial cost at the beginning of the investment. Initial costs includes cost to open a broker account and maintenance cost.

























